\texttt{nuance}: Detection of planetary transits in the presence of correlated noise

### 2023-12-03 latex/paper.tex
fix: add processing times + GP acronym

\begin{abstract}
    We present \nuancemethod{}, an algorithm to search for planetary transits in light curves featuring correlated noise, such as instrumental signals and stellar photometric variability. To deal with these nuisance signals, a common approach consists in cleaning a light curve from correlated noise before searching for transits. However, we show that this approach, based on the prior assumption that transits are not present, strongly degrades their signals, up to the point of no detection. As this degradation depends on the correlated noise characteristics, we explore the parameter space for which transits are altered, and quantify this effect on a wide variety of cases. We show that \nuancemethod{} outperforms the detection capabilities of commonly used transit search algorithms, especially for light curves featuring correlated noise with an amplitude three times greater than the searched transit depth. We perform our tests by injecting transits on synthetic light curves, and on known TESS candidates light curves to assess the performance of our algorithm on realistic multi-planetary systems datasets. Beyond its detection efficiency, we make \nuance{} tractable, \TODO{} times faster than current alternatives based on brute force detrending, and available as a well documented open-source Python package\footnote{\href{https://github.com/lgrcia/nuance}{https://github.com/lgrcia/nuance}}.
\end{abstract}

@@

\begin{figure}[H]
    \begin{centering}
        \includegraphics[width=0.6\linewidth]{../workflows/plot_issues/figures/issue1.pdf}
        \caption{Illustration of the effect of correlated noise on a single transit SNR. A 1-hour transit signal of depth 1\% is generated on top of white noise (variance of $0.0015^2$) as part of a 24-hours observation with an exposure time of 1 minute (top). Then, in the bottom plot, correlated noise is added, generated using a Gaussian Process with a Matèrn-32 kernel of scale 1 hour and sigma of 0.2\%. The SNR on the right of each light curve is computed using \autoref{eq:snr}. Models used to simulate these data are provided in \autoref{signals_simulations}.}
        \label{fig:issue1}
    \end{centering}
\end{figure}

@@

        \caption{(Top left) Simulated observation spanning 3 days, including a transit signal with a depth of 0.8\%, duration of 0.1 day, and measurement error of 1.5\% (using the model from \cite{protopapas} described in \autoref{app_proto}). (Left) Correlated noise in the form of stellar variability with different timescales and amplitudes added to the unique transit signal. This signal is then reconstructed using (in purple) Tukey's bi-weight filter \citep{wotan} with an optimal window size of three times the transit duration, and (in red) a Gaussian Process with the same kernel used to simulate the data. In each case the variability is reconstructed, subtracted, and the transit SNR computed using \autoref{eq:snr} with the transit depth estimated as the minimum in-transit flux.}

@@

The presence of instrumental correlated noise motivated the development of systematics detrending algorithms, such as the Trend Filtering Algorithm (\textsc{TFA}, \citealt{tfa}, in its primary use case), \textsc{SysRem} (\citealt{sysrem}) or Pixel Level Decorrelation (\textsc{PLD}, \citealt{pld}; see also \textsc{Everest} from \citealt{everest1, everest2}). Most of these methods rely on the shared nature of instrumental signals among light-curves (or neighboring pixels) such that the correction applied should not degrade the transit signal and can be modeled using contemporaneous measurements (like detector's temperature, pointing error, sky background or airmass time series). In opposition, stellar variability and other astrophysical signals cannot be correlated with simultaneous measurements. This gave rise to several treatments in order to reconstruct and detrend stellar variability. One of them is physically-motivated and makes use of Gaussian processes (e.g. \citealt{k2sc}), another is empirical and makes use of filtering algorithms (\citealt{Jenkins2010}, \citealt{wotan}).\\\\
In \autoref{fig:issue2}, we simulate a transit signal on top of which we add photometric stellar variability with different amplitudes and timescales, sampled from a Gaussian Process with an SHO kernel described in \autoref{app_gp}. For each light curve, we reconstruct and detrend stellar variability in two ways: one using the widely-adopted Tukey's bi-weight filter, presented in \cite{tukey} and using the implementation from \textsf{wõtan}\footnote{\href{https://github.com/hippke/wotan}{https://github.com/hippke/wotan}} \citep{wotan}; the other using the same Gaussian Process from which the data has been sampled. We then estimate the resulting transit depth and compute the remaining transit SNR using \autoref{eq:snr}. \autoref{fig:issue2} clearly shows the effect of both detrending techniques on transits SNR, and intuitively suggests that the degradation is strongly dependant on the correlated noise characteristics encountered.\\\\

@@

Ideally, we would detect the periodic transit signal in this flux by sampling the posterior likelihood of this data to a full-fledge model including stellar variability (more generally correlated noise), instrumental systematic signals (modeled with explanatory variables or treated as correlate noise), and a periodic transit signal of period $P$, epoch $T_0$, duration $D$ and depth $\Delta$. We would then reduce the posterior likelihood to $p(\bm{f}\vert P)$, its marginalized version over all parameters except the period $P$, producing a transit search periodogram $\mathcal{Q}(P)$. However, this approach has two issues: It is highly intractable, and it may lead to multimodal distributions that are hard to interpret.
\\\\
Given a period $P$, we instead want to compute the likelihood of a periodic transit signal at the maximum likelihood parameters $\hat T_0$, $\hat D$ and $\hat \Delta$, i.e the periodogram
\begin{equation}\label{eq:periodogram}
        \mathcal{Q}(P) = p(\bm{f} \vert P, \hat T_0 ,\hat D, \hat \Delta)
\end{equation}
This is done by adopting the strategy of \cite{foreman2016}, and separate the transit search into two components: the \textit{linear search} and the periodic search. During the \textit{linear search}, the likelihood of a single non-periodic transit is computed for a grid of epochs, durations and depths. Then, the periodic search consists in combining these likelihoods to compute the likelihood of the data given a periodic transit signal for a range of periods. These combined likelihoods yield a transit-search periodogram on which the periodic transit detection is based. \textsf{nuance} differs from \cite{foreman2016} and other existing transit search algorithms as it models the covariance of the light curve with a Gaussian Process, accounting for correlated noise (especially in the form of stellar variability) while keeping the model linear and tractable. This way, \textsf{nuance} searches for transits while, at the same time, modeling correlated noise, avoiding the detrending step that degrades transit signals SNR.\\\\
We note that the approach employed by \textsf{nuance} (using the two steps proposed by \citealt{foreman2016}), shares similarities with the approach of \cite{Jenkins2010}, where a single event statistic is computed and combined into a multiple event statistics.

@@

During the \textit{linear search}, the goal is to compute the likelihood $p(\bm{f} \vert T , D, \Delta)$ of the data given a single non-periodic transit signal of epoch $T$, duration $D$ and depth $\Delta$, for a grid of epochs, durations and depths.
\\\\
To account for correlated noise, the light curve $f$ is modeled as being drawn from a Gaussian Process, like the simulated light curves used in \autoref{issues}, such that
\begin{equation*}
    \bm{f} \sim \mathcal{N}(\bm{X w}, \bm{\Sigma}),
\end{equation*}
with mean $\bm{Xw}$ (i.e. a linear model of the M explanatory variables with coefficients $\bm{w}$) and covariance $\bm{\Sigma}$. To account for the presence of a single non-periodic transit of epoch $T$ and duration $D$, this signal is computed and appended as the last column of the design matrix $\bm{X}$, using the simple transit model from \cite{protopapas} with a unitary depth (\autoref{eq:protopapas}). This way, the transit signal is part of the linear model and its depth $\Delta$ can be solved linearly. Under this assumption, the log-likelihood of the data given a single non-periodic transit is \citep{Rasmussen2005}
\begin{equation} \label{eq:linear_search_ll}
    \ln p(\bm{f} \vert I) = -\frac{1}{2}(\bm{f}-\bm{Xw})^T\bm{\Sigma}^{-1}(\bm{f}-\bm{Xw}) -  \frac{1}{2}\ln\vert\bm{\Sigma}\vert - \frac{N}{2}\ln 2\pi,
\end{equation}
where the parameters vector $\bm{w}$ and their errors $\bm{\sigma}$ are computed using the generalized least-square solution
\begin{equation}\label{eq:ls}
    \bm{w} = (\bm{X}^T\bm{\Sigma}^{-1}\bm{X})^{-1}\bm{X}^T\bm{\Sigma}^{-1}\bm{f} \hspace{0.5cm} \text{and} \hspace{0.5cm} \bm{\sigma} = (\bm{X}^T\bm{\Sigma}^{-1}\bm{X})^{-1},
\end{equation} 
with $\bm{\Sigma}$ the covariance matrix modeled using the Gaussian Process. Hence, $\ln p(\bm{f} \vert I)$ can be computed on a grid of epochs and durations, the transit depth being linearly solved for any $(T, D)$. In this equation. This \textit{linear search} leads to the set of likelihoods
\begin{equation*}
    \set{\ln\mathcal{L}_{i,j}}_{i, j} = \set{\ln p(\bm{f} \vert T_i ,D_j, \Delta_{i,j})}_{i, j},
\end{equation*}
were $\Delta_{i,j}$ is the depth linearly solved for a given $(T_i, D_j)$\footnote{The expression of $\set{\ln\mathcal{L}_{i,j}}_{i, j}$ omits the vector $\bm{w}_{i,j}$ (except its last value $\Delta_{i,j}$) as it is also linearly solved for any given value of $(T_i, D_j)$ and irrelevant in what follows.}. \autoref{fig:linear_search} shows this likelihood grid computed for a simulated dataset, using the same Gaussian Process  and design matrix $\bm{X}$ used to simulate the data.

@@

Using \autoref{eq:result}, we can now compute $\ln\mathcal{P}$ for a range of periods and build a transit search periodogram using \autoref{eq:periodogram}. But a final issue emerges, one that is fundamentally linked to our strategy. Each likelihood $p(\bm{f} \vert T, D, \Delta)$ estimated during the \textit{linear search} is computed using N measurements. Hence, combining transits in the periodic search, through $\Delta_k$, $\sigma_k$ and the product of $K$ likelihoods $\set{\mathcal{L}_k}_k$ (see \autoref{eq:result}), artificially leads to a likelihood involving $N\times K$ measurements. This lead to a normalization issue when trying to compare the joint log-likelihoods $\mathcal{P}(P)$ from one period to another, as the number of observed transits differs from one period to another. This motivates a final step to produce the transit search periodogram $\mathcal{Q}$. For any period P, instead of taking $ \mathcal{Q}(P)$ as the maximum value of $\ln\mathcal{P}$, the maximum likelihood parameters
\begin{equation}\label{eq:phi0}
    (\phi_0 ,D) = \argmax_{\phi_i, D_j} \set{\ln p(\bm{f} \vert P, \phi_i, D_j)}_{i, j}
\end{equation}
are retrieved, and $\mathcal{Q}(P)$ is defined as the SNR of the transit of period $P$, epoch $T_0 = \phi_0 P$, duration $D$ and depth $\Delta$, i.e.
\begin{equation*}
    \mathcal{Q}(P) = \frac{\Delta}{\sigma},
\end{equation*}
where $\Delta$ and $\sigma$ are obtained using \autoref{eq:sigma_delta} with the last column of $X$ containing a periodic transit signal of period $P$, epoch $T_0$, duration $D$ and depth $1$. This process and the resulting periodogram $\mathcal{Q}$ are shown in \autoref{fig:periodogram}.
\begin{figure}[H]
    \begin{centering}
        \includegraphics[width=\linewidth]{../workflows/principle/figures/principle_Q.pdf}
        \caption{For each period $P$, the joint likelihood $\mathcal{P}(P)$ is computed using \autoref{eq:result}, and the value of the maximum likelihood transit SNR retained as $\mathcal{Q}(P)$.}
        \label{fig:periodogram}
    \end{centering}
\end{figure}
The periodic transit of period $P$ with the maximum SNR, i.e. maximizing $\mathcal{Q}$, is adopted as the best candidate, basing the confidence in this signal through its SNR. The parameters of this transit are the period $P$, the epoch $T_0 = \phi_0 P$ and duration $D$ (\autoref{eq:phi0}), and the depth $\Delta$ with error $\sigma$ (given by \autoref{eq:sigma_delta}).

\subsection{An open-source python package}
Methods presented in this paper are made accessible through the \textsf{nuance} open-source Python package, hosted on Github\footnote{\href{https://github.com/lgrcia/nuance}{https://github.com/lgrcia/nuance}} and released on the Python Package Index (PyPI)\footlink{https://pypi.org/project/nuance/}. 
\\\\
To instantiate a search, a user can start by creating a \textsf{Nuance} object with
\begin{lstlisting}[language=Python]
from nuance import Nuance

@@

t0s = time.copy()
# a range of 10 durations
Ds = np.linspace(0.01, 0.2, 10)
nu.linear_search(t0s, Ds)
\end{lstlisting}
Finally, the periodic search is run with
\begin{lstlisting}[language=Python]
# a wide range of periods
periods = np.linspace(0.1, 5, 2000)
search = nu.periodic_search(periods)
\end{lstlisting}
From this \texttt{search} object, the best transiting candidate parameters can be computed (\textsf{search.best}), or the $\mathcal{Q}$ periodogram retrieved (\textsf{search.Q\_snr}), together with valuable information about the transit search. The \textsf{Nuance} object also provides methods to perform transit search on light curves from multi-planetary systems, the advantage of \textsf{nuance} being that the \textit{linear search} only needs to be performed ones, and reused for the search of several transiting candidates. An extensive and maintained online documentation is provided at \href{https://nuance.readthedocs.io}{\texttt{nuance.readthedocs.io}}.

@@

To start testing \textsf{nuance} against existing methods, a simple light curve is simulated, consisting in pure white noise with a standard deviation of $5\times 10^{-4}$ (relative flux), observed for a duration of 6 days with a cadence of 2 minutes.
In this signal, box-shaped transits are injected with periods randomly sampled from 0.3 to 2.5 days, a unique transit duration of 50 minutes, and depths randomly sampled from $2.6 \times 10^{-4}$ to $6.8 \times 10^{-4}$ (using the simple model from \cite{protopapas} presented in \autoref{app_proto} with $c=500$). By design, these injected signals have an SNR ranging from 5 to 30 (with $\sigma_r = 0$ in \autoref{eq:snr}). For each light curve, transits are searched using two different methods: \textsf{nuance}, using its implementation from the Python package described in the previous section; and \texttt{BLS}, the box-least-square algorithm \citep{bls}, using the \textsf{astropy} \textsf{BoxLeastSquares} implementation\footlink{https://docs.astropy.org/en/stable/api/astropy.timeseries.BoxLeastSquares.html}. For \nuance{}\\\\
For both methods, 3000 trial periods from 0.2 to 2.6 days are searched, with a single trial duration fixed to the unique known duration of 50 minutes. A transit signal is considered detected if the absolute difference between the injected and the recovered period is less than 0.01 day. To ease the detection criteria, transit periods recovered at half or twice the injected period (aka \textit{aliases}) are considered as being detected. For this reason, detected transit epochs are not considered. Results from this \textit{injection-recovery} procedure are shown in \autoref{fig:control}.

\begin{figure}[H]
    \begin{centering}
        \includegraphics[width=\linewidth]{../workflows/control_test_bls/figures/control_test.pdf}
        \caption{Binned statistics of the injection-recovery of 4000 transit signals in a flat light curve with only white noise (whose characteristics are described in the main text), using \texttt{BLS} and \textsf{nuance}. The color scale indicates the recovery of transits in the corresponding (period, depth) parameters space, white for a full recovery and black for no detection.}
        \label{fig:control}
    \end{centering}
\end{figure}

@@

\newpage
\section{Results}\label{results}
\subsection{Comparison with \texttt{biweight+BLS} on simulated light curves}\label{simu}
\autoref{fig:snr_detrend} shows that \textsf{nuance}'s full-fledged modeling capabilities may not always be necessary and may only be beneficial for certain noise characteristics, relative to the searched transit parameters. Here, we evaluate the performance of \textsf{nuance} in the relative parameter space $(\tau, \delta)$ described in \autoref{eq:relative_params} and see when its specific treatment of correlated noise in the transit search becomes necessary.\\\\
We compare \textsf{nuance} to the approach that involves removing stellar variability from the light curves before performing the search on a detrended dataset. Similar to \autoref{detrending_effect}, an optimal bi-weight filter implemented in the \textsf{wõtan} Python package\footlink{https://github.com/hippke/wotan} is used with a window size three times that of the injected transit duration. A transit search on the detrended light curve is then performed using the \texttt{BLS} algorithm, like in \autoref{control} using the \textsf{astropy} \textsf{BoxLeastSquares} implementation\footlink{https://docs.astropy.org/en/stable/api/astropy.timeseries.BoxLeastSquares.html}. This strategy is widely used in the community and is denoted \texttt{biweight+BLS}. In what follows, the transit detection criteria are the same as the ones used in \autoref{control}, i.e. that a transit is considered recovered if the absolute difference between the injected and recovered period is less than 0.01 days, including periods found at half or twice the injected ones (aliases) and ignoring the value of the recovered transit epoch.

@@

In order to further validate nuance on a realistic dataset, we focus this section on the multi-sector TESS light curves of TOI-540, and the search for its earth-like companion TOI-540 b. We download the 2 min cadence SPOC PDCSAP light curves of TOI-540 observed in 5 sectors (4, 5, 6, 31 and 32). Like in the previous section, we use a Lomb-Scargle periodogram and identify the 0.72 days rotation period of the star, that we use as an initial value to optimize the kernel described in \autoref{rotation_kernel} on each sector independently. Here again, we employ an upper-sigma-clipping to mask flares out of the data. The resulting light curve for sector 4 and its mean model is shown in \autoref{fig:toi540_clean}.

\begin{figure}[H]
    \begin{centering}
        \includegraphics[width=\linewidth]{../workflows/comparison_toi/figures/TOI 540/4.pdf}
        \caption{Sector 4 light curve of TOI 540. The clean signal (gray points) has been mask masked for flares (black points), and the black line corresponds to the mean of the Gaussian Process model.}
        \label{fig:toi540_clean}
    \end{centering}
\end{figure}

>>>>

\keywords{transit search, time series analysis, Gaussian processes}

\begin{abstract}
    We present \nuancemethod{}, an algorithm to search for planetary transits in light curves featuring correlated noise, such as instrumental signals and stellar photometric variability. To deal with these nuisance signals, a common approach consists in cleaning a light curve from correlated noise before searching for transits. However, we show that this approach, based on the prior assumption that transits are not present, strongly degrades their signals down to no detection. As this degradation depends on the correlated noise characteristics, we explore the relative parameter space for which transits are altered, and quantify this effect on a set of simulated light curves. On this synthetic dataset, we show that \nuancemethod{} largely outperforms the detection capabilities of a commonly used transit search technique, which consists in cleaning light curves with an optimal bi-weight filter before searching for transits using a box-least-square algorithm. In order to assess the performance of \nuancemethod{} on real datasets, we inject and recover transits into the TESS light curves of 438 M dwarfs, and compare our method to 5 detrending techniques followed by a box-least-square search. For transits with a duration exceeding one fifth of the stellar variability timescale, i.e. when both signals start ressembling each other, we show that our algorithm is the most perforant in 94\% of cases, leading to both the highest number of true positives and the lowest number of false positives detections. Although simultaneously searching for transits while modeling correlated noise is expected to be computationaly expensive, we make our algorithm tractable and available as the \textsf{JAX}-powered Python package \href{https://github.com/lgrcia/nuance}{\nuancecode{}}, allowing its use on distributed environments and GPU devices. Finally, we explore the prospects offered by the \nuancemethod{} formalism, and its use to advance our knowledge of both active stars and their transiting exoplanets, both using space-based and sparse ground-based observations.
\end{abstract}

@@

\begin{figure}[H]
    \begin{centering}
        \includegraphics[width=0.6\linewidth]{../workflows/plot_issues/figures/issue1.pdf}
        \caption{Illustration of the effect of correlated noise on a single transit SNR. A 1-hour transit signal of depth 1\% is generated on top of white noise (variance of $0.0015^2$) as part of a 24-hours observation with an exposure time of 1 minute (top). Then, in the bottom plot, correlated noise is added, generated using a Gaussian process (GP) with a Matèrn-32 kernel of scale 1 hour and sigma of 0.2\%. The SNR on the right of each light curve is computed using \autoref{eq:snr}. Models used to simulate these data are provided in \autoref{signals_simulations}.}
        \label{fig:issue1}
    \end{centering}
\end{figure}

@@

        \caption{(Top left) Simulated observation spanning 3 days, including a transit signal with a depth of 0.8\%, duration of 0.1 day, and measurement error of 1.5\% (using the model from \cite{protopapas} described in \autoref{app_proto}). (Left) Correlated noise in the form of stellar variability with different timescales and amplitudes added to the unique transit signal. This signal is then reconstructed using (in purple) Tukey's bi-weight filter \citep{wotan} with an optimal window size of three times the transit duration, and (in red) a GP with the same kernel used to simulate the data. In each case the variability is reconstructed, subtracted, and the transit SNR computed using \autoref{eq:snr} with the transit depth estimated as the minimum in-transit flux.}

@@

The presence of instrumental correlated noise motivated the development of systematics detrending algorithms, such as the Trend Filtering Algorithm (\textsc{TFA}, \citealt{tfa}, in its primary use case), \textsc{SysRem} (\citealt{sysrem}) or Pixel Level Decorrelation (\textsc{PLD}, \citealt{pld}; see also \textsc{Everest} from \citealt{everest1, everest2}). Most of these methods rely on the shared nature of instrumental signals among light-curves (or neighboring pixels) such that the correction applied should not degrade the transit signal and can be modeled using contemporaneous measurements (like detector's temperature, pointing error, sky background or airmass time series). In opposition, stellar variability and other astrophysical signals cannot be correlated with simultaneous measurements. This gave rise to several treatments in order to reconstruct and detrend stellar variability. One of them is physically-motivated and makes use of GP (e.g. \citealt{k2sc}), another is empirical and makes use of filtering algorithms (\citealt{Jenkins2010}, \citealt{wotan}).\\\\
In \autoref{fig:issue2}, we simulate a transit signal on top of which we add photometric stellar variability with different amplitudes and timescales, sampled from a GP with an SHO kernel described in \autoref{app_gp}. For each light curve, we reconstruct and detrend stellar variability in two ways: one using the widely-adopted Tukey's bi-weight filter, presented in \cite{tukey} and using the implementation from \textsf{wõtan}\footnote{\href{https://github.com/hippke/wotan}{https://github.com/hippke/wotan}} \citep{wotan}; the other using the same GP from which the data has been sampled. We then estimate the resulting transit depth and compute the remaining transit SNR using \autoref{eq:snr}. \autoref{fig:issue2} clearly shows the effect of both detrending techniques on transits SNR, and intuitively suggests that the degradation is strongly dependant on the correlated noise characteristics encountered.\\\\

@@

Ideally, we would detect the periodic transit signal in this flux by sampling the posterior likelihood of this data to a full-fledge model including stellar variability (more generally correlated noise), instrumental systematic signals (modeled with explanatory variables or treated as correlate noise), and a periodic transit signal of period $P$, epoch $T_0$, duration $D$ and depth $\Delta$. We would then reduce the posterior likelihood to $p(\bm{f}\vert P)$, its marginalized version over all parameters except the period $P$, producing a transit search periodogram $\mathcal{Q}(P)$. However, this approach has two issues: It is highly intractable, and it may lead to multimodal distributions that are hard to interpret.
\\\\
Given a period $P$, we instead want to compute the likelihood of a periodic transit signal at the maximum likelihood parameters $\hat T_0$, $\hat D$ and $\hat \Delta$, i.e the periodogram
\begin{equation}\label{eq:periodogram}
        \mathcal{Q}(P) = p(\bm{f} \vert P, \hat T_0 ,\hat D, \hat \Delta)
\end{equation}
This is done by adopting the strategy of \cite{foreman2016}, and separate the transit search into two components: the \textit{linear search} and the periodic search. During the \textit{linear search}, the likelihood of a single non-periodic transit is computed for a grid of epochs, durations and depths. Then, the periodic search consists in combining these likelihoods to compute the likelihood of the data given a periodic transit signal for a range of periods. These combined likelihoods yield a transit-search periodogram on which the periodic transit detection is based. \textsf{nuance} differs from \cite{foreman2016} and other existing transit search algorithms as it models the covariance of the light curve with a GP, accounting for correlated noise (especially in the form of stellar variability) while keeping the model linear and tractable. This way, \textsf{nuance} searches for transits while, at the same time, modeling correlated noise, avoiding the detrending step that degrades transit signals SNR.\\\\
We note that the approach employed by \textsf{nuance} (using the two steps proposed by \citealt{foreman2016}), shares similarities with the approach of \cite{Jenkins2010}, where a single event statistic is computed and combined into a multiple event statistics.

@@

During the \textit{linear search}, the goal is to compute the likelihood $p(\bm{f} \vert T , D, \Delta)$ of the data given a single non-periodic transit signal of epoch $T$, duration $D$ and depth $\Delta$, for a grid of epochs, durations and depths.
\\\\
To account for correlated noise, the light curve $f$ is modeled as being drawn from a GP, like the simulated light curves used in \autoref{issues}, such that
\begin{equation*}
    \bm{f} \sim \mathcal{N}(\bm{X w}, \bm{\Sigma}),
\end{equation*}
with mean $\bm{Xw}$ (i.e. a linear model of the M explanatory variables with coefficients $\bm{w}$) and covariance $\bm{\Sigma}$. To account for the presence of a single non-periodic transit of epoch $T$ and duration $D$, this signal is computed and appended as the last column of the design matrix $\bm{X}$, using the simple transit model from \cite{protopapas} with a unitary depth (\autoref{eq:protopapas}). This way, the transit signal is part of the linear model and its depth $\Delta$ can be solved linearly. Under this assumption, the log-likelihood of the data given a single non-periodic transit is \citep{Rasmussen2005}
\begin{equation} \label{eq:linear_search_ll}
    \ln p(\bm{f} \vert I) = -\frac{1}{2}(\bm{f}-\bm{Xw})^T\bm{\Sigma}^{-1}(\bm{f}-\bm{Xw}) -  \frac{1}{2}\ln\vert\bm{\Sigma}\vert - \frac{N}{2}\ln 2\pi,
\end{equation}
where the parameters vector $\bm{w}$ and their errors $\bm{\sigma}$ are computed using the generalized least-square solution
\begin{equation}\label{eq:ls}
    \bm{w} = (\bm{X}^T\bm{\Sigma}^{-1}\bm{X})^{-1}\bm{X}^T\bm{\Sigma}^{-1}\bm{f} \hspace{0.5cm} \text{and} \hspace{0.5cm} \bm{\sigma} = (\bm{X}^T\bm{\Sigma}^{-1}\bm{X})^{-1},
\end{equation} 
with $\bm{\Sigma}$ the covariance matrix modeled using the Gaussian Process. Hence, $\ln p(\bm{f} \vert I)$ can be computed on a grid of epochs and durations, the transit depth being linearly solved for any $(T, D)$. In this equation. This \textit{linear search} leads to the set of likelihoods
\begin{equation*}
    \set{\ln\mathcal{L}_{i,j}}_{i, j} = \set{\ln p(\bm{f} \vert T_i ,D_j, \Delta_{i,j})}_{i, j},
\end{equation*}
were $\Delta_{i,j}$ is the depth linearly solved for a given $(T_i, D_j)$\footnote{The expression of $\set{\ln\mathcal{L}_{i,j}}_{i, j}$ omits the vector $\bm{w}_{i,j}$ (except its last value $\Delta_{i,j}$) as it is also linearly solved for any given value of $(T_i, D_j)$ and irrelevant in what follows.}. \autoref{fig:linear_search} shows this likelihood grid computed for a simulated dataset, using the same Gaussian Process  and design matrix $\bm{X}$ used to simulate the data.

@@

Using \autoref{eq:result}, we can now compute $\ln\mathcal{P}$ for a range of periods and build a transit search periodogram using \autoref{eq:periodogram}. But a final issue emerges, one that is fundamentally linked to our strategy. Each likelihood $p(\bm{f} \vert T, D, \Delta)$ estimated during the \textit{linear search} is computed using N measurements. Hence, combining transits in the periodic search, through $\Delta_k$, $\sigma_k$ and the product of $K$ likelihoods $\set{\mathcal{L}_k}_k$ (see \autoref{eq:result}), artificially leads to a likelihood involving $N\times K$ measurements. This lead to a normalization issue when trying to compare the joint log-likelihoods $\mathcal{P}(P)$ from one period to another, as the number of observed transits differs from one period to another. This motivates a final step to produce the transit search periodogram $\mathcal{Q}$. For any period P, instead of taking $ \mathcal{Q}(P)$ as the maximum value of $\ln\mathcal{P}$, we compute the maximum likelihood parameters
\begin{equation}\label{eq:phi0}
    (\phi_0 ,D) = \argmax_{\phi_i, D_j} \set{\ln p(\bm{f} \vert P, \phi_i, D_j)}_{i, j}
\end{equation}
and defined $\mathcal{Q}(P)$ as being equal to the SNR of the transit of period $P$, epoch $T_0 = \phi_0 P$, duration $D$ and depth $\Delta$, i.e.
\begin{equation*}
    \mathcal{Q}(P) = \frac{\Delta}{\sigma},
\end{equation*}
where $\Delta$ and $\sigma$ are obtained using \autoref{eq:sigma_delta} with the last column of $X$ containing a periodic transit signal of period $P$, epoch $T_0$, duration $D$ and depth $1$. This process and the resulting periodogram $\mathcal{Q}$ are shown in \autoref{fig:periodogram}.
\begin{figure}[H]
    \begin{centering}
        \includegraphics[width=\linewidth]{../workflows/principle/figures/principle_Q.pdf}
        \caption{For each period $P$, the joint likelihood $\mathcal{P}(P)$ is computed using \autoref{eq:result}, and the value of the maximum likelihood transit SNR retained as $\mathcal{Q}(P)$.}
        \label{fig:periodogram}
    \end{centering}
\end{figure}
The periodic transit of period $P$ with the maximum SNR, i.e. maximizing $\mathcal{Q}$, is adopted as the best candidate, basing the confidence in this signal through its SNR. The parameters of this transit are the period $P$, the epoch $T_0 = \phi_0 P$ and duration $D$ (\autoref{eq:phi0}), and the depth $\Delta$ with error $\sigma$ (given by \autoref{eq:sigma_delta}).

\subsection{An open-source python package}
Methods presented in this paper are made accessible through the \nuancecode{} open-source Python package, hosted on Github\footnote{\href{https://github.com/lgrcia/nuance}{https://github.com/lgrcia/nuance}} and released on the Python Package Index (PyPI)\footlink{https://pypi.org/project/nuance/}. 
\\\\
To instantiate a search, a user can start by creating a \textsf{Nuance} object with
\begin{lstlisting}[language=Python]
from nuance import Nuance

@@

t0s = time.copy()
# a range of 10 durations
Ds = np.linspace(0.01, 0.2, 10)
nu.linear_search(t0s, Ds)
\end{lstlisting}
Finally, the periodic search is run with
\begin{lstlisting}[language=Python]
# a wide range of periods
periods = np.linspace(0.1, 5, 2000)
search = nu.periodic_search(periods)
\end{lstlisting}
From this \texttt{search} object, the best transiting candidate parameters can be computed (\textsf{search.best}), or the $\mathcal{Q}$ periodogram retrieved (\textsf{search.Q\_snr}), together with valuable information about the transit search. The \textsf{Nuance} object also provides methods to perform transit search on light curves from multi-planetary systems, the advantage of \nuancemethod{} being that the \textit{linear search} only needs to be performed ones and reused for the search of several transiting candidates (cf.\;\autoref{toi540}). An extensive and maintained online documentation is provided at \href{https://nuance.readthedocs.io}{\texttt{nuance.readthedocs.io}}.

@@

To start testing the \nuancemethod{} approach against existing methods, a simple light curve is simulated, consisting in pure white noise with a standard deviation of $5\times 10^{-4}$ (relative flux), observed for a duration of 6 days with a cadence of 2 minutes.
In this signal, box-shaped transits are injected with periods randomly sampled from 0.3 to 2.5 days, a unique transit duration of 50 minutes, and depths randomly sampled to lead to a transit SNR from 4 to 30.
% (i.e.\;depths from $2.6 \times 10^{-4}$ to $6.8 \times 10^{-4}$, using the simple model from \cite{protopapas} presented in \autoref{app_proto} with $c=500$). By design, these injected signals have an SNR ranging from 5 to 30 (with $\sigma_r = 0$ in \autoref{eq:snr}). 
For each light curve, transits are searched using two different methods: \textsf{nuance}, using its implementation from the Python package described in the previous section; and \textsf{BLS}, the box-least-square algorithm \citep{bls}, using the \textsf{astropy} \textsf{BoxLeastSquares} implementation\footlink{https://docs.astropy.org/en/stable/api/astropy.timeseries.BoxLeastSquares.html}.\\\\
For both methods, 3000 trial periods from 0.2 to 2.6 days are searched, with a single trial duration fixed to the unique known duration of 50 minutes. A transit signal is considered detected if the absolute difference between the injected and the recovered period is less than 0.01 day. To ease the detection criteria, orbital periods recovered at half or twice the injected ones (aka \textit{aliases}) are considered as being detected. For this reason, detected transit epochs are not considered. Results from this \textit{injection-recovery} procedure are shown in \autoref{fig:control}.

\begin{figure}[H]
    \begin{centering}
        \includegraphics[width=\linewidth]{../workflows/control_test_bls/figures/control_test.pdf}
        \caption{Binned statistics of the injection-recovery of 4000 transit signals in a flat light curve with only white noise (whose characteristics are described in the main text), using \textsf{BLS} and \textsf{nuance}. The color scale indicates the recovery of transits in the corresponding (period, depth) parameters space, white for a full recovery and black for no detection.}
        \label{fig:control}
    \end{centering}
\end{figure}

@@

\newpage
\section{Results}\label{results}
\subsection{Comparison with \texttt{biweight+BLS} on simulated light curves}\label{simu}
\autoref{fig:snr_detrend} shows that \textsf{nuance}'s full-fledged modeling capabilities may not always be necessary and may only be beneficial for certain noise characteristics, relative to the searched transit parameters. Here, we evaluate the performance of \textsf{nuance} in the relative parameter space $(\tau, \delta)$ described in \autoref{eq:relative_params} and see when its specific treatment of correlated noise in the transit search becomes necessary.\\\\
We compare \textsf{nuance} to the approach that involves removing stellar variability from light curves before performing the search on a detrended dataset. Similarly to \autoref{detrending_effect}, an optimal bi-weight filter implemented in the \textsf{wõtan} Python package\footlink{https://github.com/hippke/wotan} is used with a window size three times that of the injected transit duration. A transit search on the detrended light curve is then performed using the \texttt{BLS} algorithm, like in \autoref{control} using the \textsf{astropy} \textsf{BoxLeastSquares} implementation\footlink{https://docs.astropy.org/en/stable/api/astropy.timeseries.BoxLeastSquares.html}. This strategy is widely used in the community and is denoted \texttt{biweight+BLS}. In what follows, the transit detection criteria are the same as the ones used in \autoref{control}, i.e. that a transit is considered recovered if the absolute difference between the injected and recovered period is less than 0.01 days, including periods found at half or twice the injected ones (aliases) and ignoring the value of the recovered transit epoch.

@@

In order to further validate nuance on a realistic dataset, we focus this section on the multi-sector TESS light curves of TOI-540, and the search for its earth-like companion TOI-540 b. We download the 2 min cadence SPOC PDCSAP light curves of TOI-540 observed in 5 sectors (4, 5, 6, 31 and 32). Like in the previous section, we use a Lomb-Scargle periodogram and identify the 0.72 days rotation period of the star, that we use as an initial value to optimize the kernel described in \autoref{rotation_kernel} on each sector independently. Here again, we employ an upper-sigma-clipping to mask flares out of the data. The resulting light curve for sector 4 and its mean model is shown in \autoref{fig:toi540_clean}.
\begin{figure}[H]
    \begin{centering}
        \includegraphics[width=\linewidth]{../workflows/comparison_toi/figures/TOI 540/4.pdf}
        \caption{Sector 4 light curve of TOI 540. The clean signal (gray points) has been mask masked for flares (black points), and the black line corresponds to the mean of the Gaussian Process model.}
        \label{fig:toi540_clean}
    \end{centering}
\end{figure}
